---
title: "CpG Methylation Analysis Report"
author: "Your Name"
last updated : "`r Sys.Date()`"
output: md_document
knitr:
  opts_chunk:
    echo: true 
    warning: true
---
### Data
- 'Beaudin' PMID 27666010 - DEGs comparing FL versus adult HSCs - positive FC values are enriched in FL, negative values in BM.
- 'Tan' PMID 31851943 - DEGs comparing adult versus FL HSCs - positive FC values are enriched in ADULT, negative in FL (the other way around).
- The ATAC-seq datasets are from PMID 35584393 - these are pseudobulk of the scATAC-seq from this paper. They are MACS readout files - one with peaks enriched in FL vs adult HSCs, the other adult vs FL HSCs. As a sanity check, there are more accessible regions in FL compared to adult HSCs as we would expect.


### Before Outlier Filtering
```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(dplyr)
library(data.table)

df=fread("cpg_prop_by_group.csv")
df[,prop := CpG/(CpG+uCpG)]

knitr::kable(df, caption = "CpG and uCpG Counts with Methylation Proportions")
ggplot(df, aes(x = group, y = prop, fill = group)) +
  geom_boxplot(width = 0.4, alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Methylation Proportion by Group",
       y = "Methylation (%)", x = "Group") +
  theme_minimal()
```

### After Outlier Filtering 

```{r outlier}
df = df[!id %in% c("Y1","E2")]
ggplot(df, aes(x = group, y = prop, fill = group)) +
  geom_boxplot(width = 0.4, alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Methylation Proportion by Group", y = "Methylation (%)", x = "Group") +
  theme_minimal()
glm_fit <- glm(cbind(CpG, uCpG) ~ group, data = df, family = binomial)
summary(glm_fit)

cat("Null deviance:", round(glm_fit$null.deviance), "on", glm_fit$df.null, "degrees of freedom\n")
cat("Residual deviance:", round(glm_fit$deviance), "on", glm_fit$df.residual, "degrees of freedom\n")
cat("AIC:", round(AIC(glm_fit)), "\n")

invlogit <- function(x) exp(x) / (1 + exp(x))
p_E <- invlogit(coef(glm_fit)["(Intercept)"])
p_W <- invlogit(coef(glm_fit)["(Intercept)"] + coef(glm_fit)["groupW"])
p_Y <- invlogit(coef(glm_fit)["(Intercept)"] + coef(glm_fit)["groupY"])

rate_df <- data.frame(
  Group = c("E", "W", "Y"),
  Estimated_Proportion = round(c(p_E, p_W, p_Y), 4)
)

knitr::kable(rate_df, caption = "Estimated Methylation Proportions by Group (from GLM)")
```

### Differential Sites
```{r diff-sites, message=FALSE, warning=FALSE}
source("fn.r")
pthr=0.05
input="../../bigdata/2025-08-01/filtered_3rep_3x_per_group.csv.gz"
input_sig="sig.csv";
tt=fread(input)
tt[,`:=`(start = as.integer(start), end = as.integer(end))]

fwrite(tt[,.(chrom,start,end,cE)],"EM_cE.bedGraph.gz",sep="\t",col.names=F)
fwrite(tt[,.(chrom,start,end,tE)],"EM_tE.bedGraph.gz",sep="\t",col.names=F)
fwrite(tt[,.(chrom,start,end,cW)],"EM_cW.bedGraph.gz",sep="\t",col.names=F)
fwrite(tt[,.(chrom,start,end,tW)],"EM_tW.bedGraph.gz",sep="\t",col.names=F)

if( !file.exists(input_sig) ){
    res=binom(tt)
    fwrite(res[pvEW < pthr| pvWY < pthr ],file=input_sig)
}else{
    res=fread(input_sig)
}
pl_venn(res)

res[,end:=start+1]
x=merge(res[pvEW < pthr,],tt)

fwrite(x[,.(chrom,start,end,pW-pE)],file="sig_EtoW.bedGraph",sep="\t",col.names=F)
x=merge(res[pvWY < pthr,],tt)
fwrite(x[,.(chrom,start,end,pY-pW)],file="sig_WtoY.bedGraph",sep="\t",col.names=F)


```
data: [sig_diff.csv](sig.csv)


